Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2017-03-18 paper/paper-v3.tex
minor revisions up to 3.1.5

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of 
  conventional semantic meaning and general pragmatic reasoning applied
  to the context of utterance. This implies that models of cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of cultural language evolution where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a loss in expressivity otherwise incurred by the adoption of such semantics. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept the invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
thereby making a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved. Models of language evolution abound. There are
simulation-based models studying the evolution of language in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much of this work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. Instead, we would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse data to learn from. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g., in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that associates a distinct form with all possible meanings a speaker may want to
convey is maximally expressive but challenging to acquire.

@@

\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} \tb{Yes!} For our present purpose, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  
Lexica for this case would assign a Boolean truth value, either $0$ or $1$, to each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is also necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ because $\msome$ is semantically associated with both states by $\Llack$.

Pragmatic behavior of level $n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}

Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can come to {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$ in contrast to their literal counterparts. Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$. Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message. The converse reasoning pattern holds for the speaker. In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.

@@

Most evolutionary dynamics assume that the proportion of type $i$ in a population will increase
or decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is frequently associated with expressivity, i.e., the ability to successfully
communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
interpretation, the assumption is that organisms have a higher chance of survival and
reproduction if they are able to share and receive useful information via communication with
peers. Under a cultural interpretation, the picture is that agents themselves strive towards
communicative success and therefore occasionally adapt or revise their behavior to achieve a
higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

The replicator equation gives us the means to make the ensuing dynamics precise, without
necessarily committing to a biological or cultural interpretation. As above, the proportion of types in a
given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
fitness of type $i$ is its average expected communicative success, or \emph{expected
  utility} (EU), given the frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
of the time this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
  \delta(s,s') \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
\end{flalign*}
As usual for cooperative communication, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's lexicon
and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
behavior of type $i$, they observe the linguistic behavior of type $i$ and need to infer what
their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
of linguistic behavior of agents who have acquired the language from observation and inference
before as well. In the simplest case there is a single teacher and a single
learner. After sufficient training the learner becomes a teacher and produces behavior that
serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
and \citealt{tamariz+kirby:2016} for recent surveys).

>>>>

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of 
  conventional semantic meaning and general pragmatic reasoning applied
  to the context of utterance. This implies that models of the cultural evolution of meaning
  should likewise take into consideration that observable language use is a complex interaction
  of semantic representations and pragmatic use. To this end, we present a game theoretic model
  of cultural language evolution where communicative pressures work on abstract semantic
  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
  systematic transmission perturbations when linguistic knowledge is transferred from one agent
  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a loss in expressivity otherwise incurred by the adoption of such semantics. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely
veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
can be declined by saying ``I have to work.'' An influential explanation of the relation
between the literal meaning of expressions and what they are intended and interpreted to convey
is due to \citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of
mutual reasoning about rational language use. For instance, under the assumption that the
speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
reason why the speaker will not be able to accept an invitation, going beyond its literal
meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
such as the use of ability questions for polite requests (``Could you please \dots?''), or
certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.

A particularly productive and well studied class of systematic pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
thereby making a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved and why it would be so pervasive across natural languages. Models of language evolution abound. There are
simulation-based models studying the evolution of language in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much of this work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. Instead, we would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse learning data. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g., in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Importantly, even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a 
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that lexicalizes a distinct form for each possible fine-grained meaning a speaker may want to
convey is maximally expressive but challenging to acquire.

@@

\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.} \tb{Yes, I think that's a good idea. We can always cut it out again if the reviewers prefer more compression but let's explain more and cut later. The paper is relatively short at the moment.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} \tb{Yes!} For our present purpose types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  
Lexica for this case would assign a Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
semantics with only a lower-bound in $\Llack$.

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning. Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is also necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ for $\Llack$ because it semantically associates $\msome$ with both states.

Pragmatic behavior of level $n+1$ is similar to its literal counterparts but bases linguistic choice in 
interpretation or production on the expected behavior of a level-$n$ interlocutor instead of on lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{flalign}

Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts. Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$. Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message. The converse reasoning pattern holds for the speaker. In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.

@@

Most evolutionary dynamics assume that the proportion of type $i$ in a population will increase
or decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is frequently associated with expressivity, i.e., the ability to successfully
communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
interpretation the assumption is that organisms have a higher chance of survival and
reproduction if they are able to share and receive useful information via communication with
peers. Under a cultural interpretation the picture is that agents themselves strive towards
communicative success and therefore occasionally adapt or revise their behavior to achieve a
higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

The replicator equation gives us the means to make the ensuing dynamics precise, without
necessarily committing to a biological or cultural interpretation. As above, the proportion of types in a
given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
fitness of type $i$ is its average expected communicative success, or \emph{expected
  utility} (EU), given the frequencies of types in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
of the time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
  \delta(s,s') \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
\end{flalign*}
As usual for cooperative communication, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's lexicon
and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
behavior of type $i$, they observe the linguistic behavior of type $i$ and need to infer what
their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
of linguistic behavior of agents who have acquired the language from observation and inference
before as well. In the simplest case there is a single teacher and a single
learner. After sufficient training the learner becomes a teacher and produces behavior that
serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
learning alone generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
and \citealt{tamariz+kirby:2016} for recent surveys).
